Proving Logical Atomicity using Lock Invariants\thanks{Supported by the National Science Foundation via grant \#1811894.}

### 2022-05-17 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
client proof edits

\begin{figure}[htp]
\begin{subfigure}[t]{0.4\textwidth}
\begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
void* update_tree(void* t){
  a = 1;
  b = 2;
  c = 3; 
  d = 4;
  insert(t, 1, &a);
  insert(t, 2, &b);
  insert(t, 1, &c);
  insert(t, 2, &d);
  return NULL;
}
\end{lstlisting} 
\end{subfigure}\qquad
\begin{subfigure}[t]{0.5\textwidth}
 \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
int retrieve_value(treebox t){
  void* searchResult;
  int y = -1;
  do {
     searchResult = lookup (t,  2);
     if (searchResult != NULL) {
        y = *((int *) searchResult);
     }
  } while (y != 4);
   searchResult = lookup (t,  1);
   return *((int *) searchResult);
}
\end{lstlisting}
\end{subfigure}
\caption{Client code}
\label{bst-conc-client}
\end{figure}

To demonstrate that the complexities of our lock-based proofs are hidden from prospective users, we verified a simple client program for the binary search tree, shown in figure \ref{bst-conc-client}. A producer thread runs the \lstinline{update_tree} function to populate a tree, while a consumer thread runs \lstinline{retrieve_value} to wait until key 2 is associated with a pointer to the value 4, and then returns the value associated with key 1. This is a data-structure-level version of the message-passing idiom, and demonstrates one of the advantages of atomic specifications: we can use them to synchronize threads and pass ownership of resources via data structure operations. The invariant for the BST \lstinline{t} is a simple state machine with three states, depending on the contents of the abstract map $m$: 1) $m(2) = \texttt{NULL}$, 2) $m(2) \mapsto i$ for some integer $i \neq 4$; and 3) $m(2) \mapsto 4 \land m(1) \mapsto 3$. In this final state, the invariant also holds the resources to be transferred from the producer to the consumer. We also include ghost state to reflect the fact that there is exactly one producer thread (so the producer always knows exactly which values are in the tree) and one consumer thread (so it can retrieve the resources without worrying about interference). The proof of the client uses only the atomic specifications of the BST operations, and is not affected by the style of lock specification.

>>>>

\begin{figure}[htp]
\begin{subfigure}[t]{0.4\textwidth}
\begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
void* update_tree(void* t){
  a = 1;
  b = 2;
  c = 3; 
  d = 4;
  insert(t, 1, &a);
  insert(t, 2, &b);
  insert(t, 1, &c);
  insert(t, 2, &d);
  return NULL;
}
\end{lstlisting} 
\end{subfigure}\qquad
\begin{subfigure}[t]{0.55\textwidth}
 \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
int retrieve_value(treebox t){
  int y = -1;
  do {
     searchResult = lookup (t,  2);
     if (searchResult != NULL) {
        y = *((int *) searchResult);
     }
  } while (y != 4);
  void* searchResult = lookup (t,  1);
  return *((int *) searchResult);
}
\end{lstlisting}
\end{subfigure}
\caption{Client code}
\label{bst-conc-client}
\end{figure}

To demonstrate that the complexities of our lock-based proofs are hidden from prospective users, we verified a simple client program for the binary search tree, shown in figure \ref{bst-conc-client}. A producer thread running the \lstinline{update_tree} function populates a tree, while a consumer thread running \lstinline{retrieve_value} waits until key 2 is associated with a pointer to the value 4, and then returns the value associated with key 1. This is a data-structure-level version of the message-passing idiom, and demonstrates one of the advantages of atomic specifications: we can use them to synchronize threads and pass ownership of resources via data structure operations. The invariant for the BST \lstinline{t} is a simple state machine with three states, depending on the contents of the abstract map $m$: 1) $m(2) = \texttt{NULL}$, 2) $m(2) \mapsto i$ for some integer $i \neq 4$; and 3) $m(2) \mapsto 4 \land m(1) \mapsto 3$. In this final state, the invariant also holds the resources to be transferred from the producer to the consumer. We also include ghost state to reflect the fact that there is exactly one producer thread (so the producer always knows exactly which values are in the tree) and one consumer thread (so it can retrieve the resources without worrying about interference). The proof of the client uses only the atomic specifications of the BST operations, and is not affected by the style of lock specification.
